Finch: Sparse and Structured Array Programming with Control Flow

### 2024-03-22 sections/tensors.tex
figure changes

To build our bridge, we embrace a set of abstractions: level formats/Fiber Trees, iteration context dependent instantiation of iterators, and tensor life cycles.
%
Our first abstraction mostly already exists in the literature: a manner of specifying a data structure for a multi-dimensional tensor out of data structures for single dimensional tensors~\cite{sze2017efficient,chou2022compilation, chou2018format}.
%
We recapitulate the essential details here.
%
Our next two abstractions adds to to the first by providing a mechanism to use data structures generated by the first abstraction in a greater variety of contexts while maintaining per-dimension encapsulation of array data structures.
%
Previous efforts to compile a greater variety of sparse array programs left these bridges untouched ~\cite{henry_compilation_2021, won2023unified, senanayake2020sparse}.
%
We introduce an interface to instatiate iterators in a variety of contexts in our programs and we introduce the lifecycle interface to manage when we read and write to multi-dimensional iterators.
%
These interfaces add to the level abstraction, expanding the types of data that they can express via mapping to looplets and expanding the contexts in which they can be used.

@@

Instead of storing the data for each subfiber separately, most sparse tensor
formats such as CSR, DCSR, and COO usually store the data for all fibers in a
level contiguously. In this way, we can think of a level as a bulk allocator for
fibers. Continuing the analogy, we can think of each fiber as being
disambiguated by a \textbf{position}, or an index into the bulk pool of
subfibers. The mapping $f$ from indices to subfibers is thus a mapping from an
index and a position in a level to a subposition in a sublevel.
Figure~\ref{fig:help} shows a simple example of a level as a fiber allocator.

Our view of a level as a fiber allocator implies an allocation function
$assemble(tns, pos_start:pos_stop)$, which allocates fibers at positions
$pos_start:pos_stop$ in the level. We don't specify a de-allocation function,
instead relying on initialization to reset the fiber if it needs to be reused.

@@

\paragraph{$declare(lvl, init, dims...)$} Declares the level to hold subtensors of size $dims$ and an initial value of $init$.

>>>>

To build our bridge, we embrace a set of abstractions: level formats/Fiber Trees, iteration context dependent instantiation of iterators, and tensor life cycles.
%
Our first abstraction mostly already exists in the literature: a manner of specifying a data structure for a multi-dimensional tensor out of data structures for single dimensional tensors~\cite{sze2017efficient,chou2022compilation, chou2018format}.
%
We recapitulate the essential details here.
%
Our next two abstractions add to to the first by providing a mechanism to use data structures generated by the first abstraction in a greater variety of contexts while maintaining per-dimension encapsulation of array data structures.
%
We introduce an interface to instatiate iterators in a variety of contexts in our programs and we introduce the lifecycle interface to manage when we read and write to multi-dimensional iterators.
%
These interfaces add to the level abstraction, expanding the types of data that they can express via mapping to looplets and expanding the contexts in which they can be used.
%
Previous efforts to compile a greater variety of sparse array programs left these bridges untouched ~\cite{henry_compilation_2021, won2023unified, senanayake2020sparse}.

@@

Instead of storing the data for each subfiber separately, most sparse tensor
formats such as CSR, DCSR, and COO usually store the data for all fibers in a
level contiguously. In this way, we can think of a level as a bulk allocator for
fibers. Continuing the analogy, we can think of each fiber as being
disambiguated by a \textbf{position}, or an index into the bulk pool of
subfibers. The mapping $f$ from indices to subfibers is thus a mapping from an
index and a position in a level to a subposition in a sublevel.
Figure~\ref{fig:levelsvsfibers} shows a simple example of a level as a pool of fibers.

When we need to refer to a particular fiber at position $p$ in the level $l$, we
may write $fiber(l, p)$. Note that the formation of fibers from levels is lazy,
and the data underlying each fiber is managed entirely by the level, so the
level may choose to overlap the storage between different fibers. Thus, the only
unique data associated with $fiber(l, p)$ is the position $p$.

\begin{figure}
    \centering
    \includegraphics[width=0.45\linewidth]{LevelsVsFibers-matrix.png}\hfill%
    \includegraphics[width=0.5\linewidth]{LevelsVsFibers-tensor.png}
    \caption{Levels and fiber tree representations of a sparse matrix and a sparse tensor. On left, a matrix is represented in a fibertree corresponding to CSC format, with a dense outer level and a sparse inner level. On right, a tensor is represented in a fibertree with two sparse outer levels, and a dense inner level. Note that the element levels in this case form the leaves of the tree.}
    \label{fig:levelsvsfibers}
\end{figure}

@@

Our view of a level as a fiber allocator implies an allocation function
$assemble(tns, pos_{start}:pos_{stop})$, which allocates fibers at positions
$pos_{start}:pos_{stop}$ in the level. We don't specify a de-allocation function,
instead relying on initialization to reset the fiber if it needs to be reused.

\paragraph{$declare(lvl, init, dims...)$} Declares the level to hold subtensors of size $dims$ and an initial value of $init$. 
\paragraph{$declare(lvl, init, dims...)$} Declares the level to hold subtensors of size $dims$ and an initial value of $init$. 
\paragraph{$freeze(lvl, init, dims...)$} Declares the level to hold subtensors of size $dims$ and an initial value of $init$.
